Reducing Linearizability of Priority Queues and More Data Structures to State Reachability

### 2017-04-20 co-regular-EPQ.tex
first part of co-regular

We show that the set of executions for which some projection fails the test $\textit{Check-PQ-Conc-NonRec}$ can be characterized using a set of register automata, modulo a value renaming. The possibility of renaming values (which is complete for checking data independent implementations) allows to simplify the reasoning about projections. Thus, we assume that all the operations which are not in the projection failing this test use the same distinguished value $\top$, different from those used in the projection. Then, it is enough to find an automata characterization for the set of executions $e$ for which $\textit{Check-PQ-Conc-NonRec}$ fails, or equivalently, for which one 
%a projection $e'$ fails $\textit{Check-PQ-Conc-NonRec}$ when one 
of the following three formulas is false:
\begin{align*}
\hspace{-5mm}
\Gamma(e) := \mathsf{Has\text{-}}\Gamma(e) \Rightarrow \exists \alpha.\ \Gamma\mathsf{\text{-}Conc}(e,\alpha)\mbox{ with $\Gamma\in \{\mathsf{EmptyRemove}$, $\mathsf{UnmatchedMaxPriority}$, $\mathsf{MatchedMaxPriority}\}$}
\end{align*}
Extending the automaton corresponding to some $\Gamma$ with self-loops that allow any operation with parameter $\top$ results in an automaton satisfying the following property called \emph{$\Gamma$-completeness}.

@@

$A \cap \mathcal{I} \neq \emptyset$ if and only if there exists $ e \in \mathcal{I}$ and $e' \in \textit{proj}(e)$ such that $\Gamma(e')$ doesn't hold.
\end{definition}

@@

W.l.o.g., we assume that every implementation $\mathcal{I}$ behaves correctly, i.e., as a FIFO queue, when only values with the same priority are observed. More precisely, we assume that for every execution $e\in\mathcal{I}$ and every priority $p\in\mathbb{P}$, the projection of $e$ to values with priority $p$ is linearizable. This property can be checked separately using the techniques introduced for regular FIFO queues in~\cite{DBLP:conf/icalp/BouajjaniEEH15} (see Appendix~\ref{} for more details). Note that this assumption excludes some obvious violations, such as a $\textit{rm}(a)$ operation happens before a $\textit{put}(a,p)$ operation, for some $p$.

Also, we consider $\Gamma$-complete automata for $\Gamma\in \{\mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$, recognizing executions where any two occurring priorities are comparable (w.r.t. $\prec$). This is possible because any data-differentiated execution for which $\Gamma(e)$ is false has such a projection. Formally, given a data-differentiated execution $e$ and $p$ a maximal priority in $e$, $e\vert_{\preceq p}$ is the projection of $e$ to the set of values with priorities comparable to $p$. Then,

\begin{restatable}{lemma}{priExecutionIsEnough}
\label{lemma:pri execution is enough}
Let $\Gamma\in \{\mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$ and $e$ be a data-differentiated execution. Then, $\Gamma(e)$ is false if and only if $\Gamma(e\vert_{\preceq p})$ is false for some maximal priority $p$ in $e$.
\end{restatable}
\begin {proof} (Sketch)
TODO UPDATE THE FOLLOWING
To prove the $\textit{only if}$ direction, given $e \sqsubseteq l = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w \in \mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,\alpha)$. Since there is no restriction to $p$-incomparable operations in $u$, $v$ and $w$, we can safely erase all $p$-incomparable values. Similar case holds for $\Gamma=\mathsf{UnmatchedMaxPriority}$.

@@

The formula $\mathsf{MatchedMaxPriority}$ is false for some execution $e$ when all the values of maximal priorities in $e$ are removed and there exists a value $x$ of maximal priority $p$ s.t. $\mathsf{MatchedMaxPriority\text{-}Conc}(e,x)$ is false. We consider two cases depending on whether $x$ is the only value with priority $p$. We denote by $\mathsf{MatchedMaxPriority}^{>}$ the strengthening of $\mathsf{MatchedMaxPriority}$ with this condition (i.e., all the values other than $x$ have a priority strictly smaller than $p$) and $\mathsf{MatchedMaxPriority}^{=}$ for the strengthening of the same formula with the negation of this condition (i.e., there exists another value $y$ with priority $p$).

Register automata is a kind of finite automata with registers used to deal with infinite data domain, where register can be assigned values and do simple operations. In this paper we use a restricted kind of register automata, which contains only one register, whose value is obtained once by guessing and never changed. Our register machine contains a register $r$. Its transition labels are $r=*,\textit{call}(\textit{put},d,g), \textit{ret}(\textit{put},d,g),$ $\textit{call}(\textit{rm},d),\textit{ret}(\textit{rm},d)$, where $d \in \mathbb{D} \cup \{ \textit{empty} \}$, and $g$ and $\top$ are guards. The first transition from initial state is a $r=*$ transition, and is used to assign an arbitrary value to $r$. A guard is a predicate chosen from $\{=r,< r,\top \}$, which represents equals the value of $r$, less than the value of $r$, and always holds, respectively. Given an execution $e = \alpha_1 \cdots \alpha_k$ of priority queue, register automata $\mathcal{A}$ accepts $e$, if there exists value $d_r\in \mathbb{D}$ and transitions $q_0 \xrightarrow{\beta_1} q_1 \ldots \xrightarrow{\beta_k} q_k$ of $\mathcal{A}$ from a initial state to a final state, such that for each $i$

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] If $\alpha_i = \textit{call}(\textit{put},a,p)$, then $\beta_i= \textit{call}(\textit{put},a,g)$ and $g$ holds with $p$ and $d_r$. Similarly for $\textit{ret}(\textit{put},a,p)$.

\item[-] Else, $\alpha_i = \beta_i$.
\end{itemize}

@@

To facilitate the proof of co-Regular of $\mathsf{MatchedMaxPriority}$, we separate $\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)$ into the following two sub-predicates, where the first one represents that $x$ is the only value with priority $p$ in $e$ , and the second one represents that there are more than one value with priority $p$ in $e$. Here $\textit{occurOnce}(p,l)$ holds if there is only one value with priority $p$ in $l$.

{\small
\begin{align*}
\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Seq}(e,x)=\mathsf{true} & \mbox{ iff  $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(e,x)=\mathsf{true}$,} \\
&\hspace{4mm}\mbox{$\textit{occurOnce}(p,e)$ where $p$ is the priority of $x$} \\
\mathsf{MatchedMaxPriority}^{=}\mathsf{\text{-}Seq}(e,x)=\mathsf{true} & \mbox{ iff  $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(e,x)=\mathsf{true}$,} \\
&\hspace{4mm}\mbox{$\neg\textit{occurOnce}(p,e)$ where $p$ is the priority of $x$} \\
\end{align*}
}

$\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Conc}(e,x)$ is obtained from $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Conc}(e,x)$ by using $\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Seq}(e,x)$ instead of $\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)$. And similarly for $\mathsf{MatchedMaxPriority}^{=}\mathsf{\text{-}Conc}(e,x)$.

We can also define two conditions, where $\textit{unmatched-priorities}(l)$ is the set of priorities occurring in $\textit{put}$ operations of $e$ for which there is $\textit{rm}$ operation removing the same value:

{\small
\begin{align*}
\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Condition}(e)=\mathsf{true} & \mbox{ iff $p\in \textit{matched-priorities}(e)$ for a maximal priority} \\
&\hspace{4mm}\mbox{$p\in priorities(e)$, $\textit{occurOnce}(p,e)$} \\
\mathsf{MatchedMaxPriority}^{=}\mathsf{\text{-}Condition}(e)=\mathsf{true} & \mbox{ iff $p\in \textit{matched-priorities}(e)$ for a maximal priority} \\
&\hspace{4mm}\mbox{$p\in priorities(e)$, $\neg\textit{occurOnce}(p,e)$} 
\end{align*}}

In this subsection, we briefly introduce the idea for proving co-regular of $\mathsf{MatchedMaxPriority}^{>}$. The proof of this subsection can be found in Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}. The notion of left-right constraint used for priority queue is inspired by left-right constraint of queue \cite{Bouajjani:2015}.

Given a data-differentiated $\_$-execution $e$ such that $\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Condition}(e)$ holds, Lemma \ref{lemma:automata for extended priority queue with single priority} is not enough for ensure linearizable w.r.t $\mathsf{MatchedMaxPriority}^{>}$. This is because that it is possible that $\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Conc}(e,x)$ does not holds because of interaction between actions of multiple priorities. One such example is shown in \figurename~\ref{fig:introduce gap for EPQ1Lar}. We call the time interval from $\textit{ret}(\textit{put},x)$ to $\textit{cal}(rm,x)$, or from $\textit{ret}(\textit{put},x)$ when $\textit{cal}(rm,x)$ does not exist, the interval of value $x$. In \figurename~\ref{fig:introduce gap for EPQ1Lar}, we draw the interval of each value by dashed line. Here we assume that $p_1 \prec p_4$, $p_2 \prec p_4$ and $p_3 \prec p_4$. We can not find a sequence $s$ where $e \sqsubseteq s$ and $\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Seq}(s,b)$ holds, since each time point from $\textit{cal}(\textit{rm},b)$ to $\textit{ret}(\textit{rm},b)$ is in interval of some value with smaller priority, and $\textit{rm}(b)$ can not happen in interval of value with smaller priority.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.4 \textwidth]{figures/PIC-HIS-INTRO-GAP-EPQ1L.pdf}
%\vspace{-10pt}
  \caption{An execution that does not linearizable w.r.t $\mathsf{MatchedMaxPriority}^{>}$}
  \label{fig:introduce gap for EPQ1Lar}
\end{figure}

@@

>>>>

We show that the set of executions for which some projection fails the test $\textit{Check-PQ-Conc-NonRec}$ can be characterized using a set of register automata, modulo a value renaming. The possibility of renaming values (which is complete for checking data independent implementations) allows to simplify the reasoning about projections. Thus, we assume that all the operations which are not in the projection failing this test use the same distinguished value $\top$, different from those used in the projection. Then, it is enough to find an automata characterization for the set of executions $e$ for which $\textit{Check-PQ-Conc-NonRec}$ fails, or equivalently, for which one 
%a projection $e'$ fails $\textit{Check-PQ-Conc-NonRec}$ when one 
of the following three formulas is false:
\begin{align*}
\hspace{-5mm}
\Gamma(e) := \mathsf{Has\text{-}}\Gamma(e) \Rightarrow \exists \alpha.\ \Gamma\mathsf{\text{-}Conc}(e,\alpha)\mbox{ with $\Gamma\in \{\mathsf{EmptyRemove}$, $\mathsf{UnmatchedMaxPriority}$, $\mathsf{MatchedMaxPriority}\}$}
\end{align*}
Intuitively, $\Gamma(e)$ states that $e$ is linearizable w.r.t. the set of sequential executions described by $\Gamma\mathsf{\text{-}Seq}$ (provided that $\mathsf{Has\text{-}}\Gamma(e)$ holds). Therefore, by an abuse of terminology, an execution $e$ satisfying $\Gamma(e)$ is called \emph{$\Gamma$-linearizable}.
Extending the automaton characterizing executions which are not $\Gamma$-linearizable, with self-loops that allow any operation with parameter $\top$ results in an automaton satisfying the following property called \emph{$\Gamma$-completeness}.

@@

$A \cap \mathcal{I} \neq \emptyset$ if and only if there exists $ e \in \mathcal{I}$ and $e' \in \textit{proj}(e)$ such that $e'$ is not $\Gamma$-linearizable.
\end{definition}

@@

W.l.o.g., we assume that every implementation $\mathcal{I}$ behaves correctly, i.e., as a FIFO queue, when only values with the same priority are observed. More precisely, we assume that for every execution $e\in\mathcal{I}$ and every priority $p\in\mathbb{P}$, the projection of $e$ to values with priority $p$ is linearizable (w.r.t. $\seqPQ$). This property can be checked separately using the techniques introduced for regular FIFO queues in~\cite{DBLP:conf/icalp/BouajjaniEEH15} (see Appendix~\ref{} for more details). Note that this assumption excludes some obvious violations, such as a $\textit{rm}(a)$ operation happens before a $\textit{put}(a,p)$ operation, for some $p$.

Also, we consider $\Gamma$-complete automata for $\Gamma\in \{\mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$, recognizing executions where any two occurring priorities are comparable (w.r.t. $\prec$). This is possible because any data-differentiated execution which is not $\Gamma$-linearizable has such a projection. Formally, given a data-differentiated execution $e$ and $p$ a maximal priority in $e$, $e\vert_{\preceq p}$ is the projection of $e$ to the set of values with priorities comparable to $p$. Then,

\begin{restatable}{lemma}{priExecutionIsEnough}
\label{lemma:pri execution is enough}
Let $\Gamma\in \{\mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$ and $e$ be a data-differentiated execution. Then, $e$ is not $\Gamma$-linearizable if and only if $e\vert_{\preceq p}$ is not $\Gamma$-linearizable, for some maximal priority $p$ in $e$.
\end{restatable}
\begin {proof} (Sketch)
TODO UPDATE THE FOLLOWING
To prove the $\textit{only if}$ direction, given $e \sqsubseteq l = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w \in \mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,\alpha)$. Since there is no restriction to $p$-incomparable operations in $u$, $v$ and $w$, we can safely erase all $p$-incomparable values. Similar case holds for $\Gamma=\mathsf{UnmatchedMaxPriority}$.

@@

A differentiated execution $e$ is not $\mathsf{MatchedMaxPriority}$-linearizable when all the $\textit{put}$ operations in $e$ using the maximal priority $p$ are matched, and $e$ is not linearizable w.r.t. the set of sequential executions satisfying $\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)$ for some value $x$ of priority $p$. We consider two cases depending on whether $e$ contains exactly one value with priority $p$ or at least two values. We denote by $\mathsf{MatchedMaxPriority}^{>}$ the strengthening of $\mathsf{MatchedMaxPriority}$ with the condition that all the values other than $x$ have a priority strictly smaller than $p$ (corresponding to the first case), and by $\mathsf{MatchedMaxPriority}^{=}$ the strengthening of the same formula with the negation of this condition (corresponding to the second case). We use particular instances of register automata~\cite{finite-memory, the paper I found} whose states include only two registers, one for storing a priority guessed at the initial state, and one for storing the priority of the current action in the execution. The transitions can check equality or the order relation $\prec$ between the values stored in the two registers. Instead of formalizing the full class of register automata, we consider a simpler class which suffices our needs. More precisely, we consider a class of labeled transition systems whose states consist of a finite control part and a register $r$ interpreted to elements of $\mathbb{P}$. The transition labels can be one of the following: 
\begin{itemize}
	\item $r=*$ for storing an arbitrary value to $r$, 
	\item $\textit{call}(\textit{rm},a)$ and $\textit{ret}(\textit{rm},a)$ for reading call/return actions of a remove, 
	\item $\textit{call}(\textit{put},d,g)$ where $g\in\{=r,\prec r,true\}$ is a guard, for reading a call action $\textit{call}(\textit{put},d,p)$ of a put and checking whether $p$ is either equal to or smaller than the value stored in $r$, or arbitrary,
	\item $\textit{ret}(\textit{put},d,true)$ for reading a return action $\textit{ret}(\textit{put},d,p)$ for some $p$.
\end{itemize}
The set of words accepted by such a transition system can be defined as usual.

%$\textit{call}(\textit{put},d,g)$ $\textit{ret}(\textit{put},d,g),$ $\textit{call}(\textit{rm},d),\textit{ret}(\textit{rm},d)$, where $d \in \mathbb{D} \cup \{ \textit{empty} \}$, and $g$ and $\top$ are guards
%
% which are complete for each case. Since we don't require the full generality of register automata, 
%
%
%We define $\Gamma$-complete automata for both cases and their union will give a $\mathsf{MatchedMaxPriority}$-complete automaton
%
%
%
%Register automata is a kind of finite automata with registers used to deal with infinite data domain, where register can be assigned values and do simple operations. In this paper we use a restricted kind of register automata, which contains only one register, whose value is obtained once by guessing and never changed. Our register machine contains a register $r$. Its transition labels are $r=*,\textit{call}(\textit{put},d,g), \textit{ret}(\textit{put},d,g),$ $\textit{call}(\textit{rm},d),\textit{ret}(\textit{rm},d)$, where $d \in \mathbb{D} \cup \{ \textit{empty} \}$, and $g$ and $\top$ are guards. The first transition from initial state is a $r=*$ transition, and is used to assign an arbitrary value to $r$. A guard is a predicate chosen from $\{=r,< r,\top \}$, which represents equals the value of $r$, less than the value of $r$, and always holds, respectively. Given an execution $e = \alpha_1 \cdots \alpha_k$ of priority queue, register automata $\mathcal{A}$ accepts $e$, if there exists value $d_r\in \mathbb{D}$ and transitions $q_0 \xrightarrow{\beta_1} q_1 \ldots \xrightarrow{\beta_k} q_k$ of $\mathcal{A}$ from a initial state to a final state, such that for each $i$
%
%\begin{itemize}
%\setlength{\itemsep}{0.5pt}
%\item[-] If $\alpha_i = \textit{call}(\textit{put},a,p)$, then $\beta_i= \textit{call}(\textit{put},a,g)$ and $g$ holds with $p$ and $d_r$. Similarly for $\textit{ret}(\textit{put},a,p)$.
%
%\item[-] Else, $\alpha_i = \beta_i$.
%\end{itemize}

@@

%To facilitate the proof of co-Regular of $\mathsf{MatchedMaxPriority}$, we separate $\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)$ into the following two sub-predicates, where the first one represents that $x$ is the only value with priority $p$ in $e$ , and the second one represents that there are more than one value with priority $p$ in $e$. Here $\textit{occurOnce}(p,l)$ holds if there is only one value with priority $p$ in $l$. 
%
%{\small
%\begin{align*}
%\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Seq}(e,x)=\mathsf{true} & \mbox{ iff  $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(e,x)=\mathsf{true}$,} \\
%&\hspace{4mm}\mbox{$\textit{occurOnce}(p,e)$ where $p$ is the priority of $x$} \\
%\mathsf{MatchedMaxPriority}^{=}\mathsf{\text{-}Seq}(e,x)=\mathsf{true} & \mbox{ iff  $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(e,x)=\mathsf{true}$,} \\
%&\hspace{4mm}\mbox{$\neg\textit{occurOnce}(p,e)$ where $p$ is the priority of $x$} \\
%\end{align*}
%}
%
%$\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Conc}(e,x)$ is obtained from $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Conc}(e,x)$ by using $\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Seq}(e,x)$ instead of $\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)$. And similarly for $\mathsf{MatchedMaxPriority}^{=}\mathsf{\text{-}Conc}(e,x)$. 
%
%We can also define two conditions, where $\textit{unmatched-priorities}(l)$ is the set of priorities occurring in $\textit{put}$ operations of $e$ for which there is $\textit{rm}$ operation removing the same value: 
%
%{\small
%\begin{align*}
%\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Condition}(e)=\mathsf{true} & \mbox{ iff $p\in \textit{matched-priorities}(e)$ for a maximal priority} \\
%&\hspace{4mm}\mbox{$p\in priorities(e)$, $\textit{occurOnce}(p,e)$} \\
%\mathsf{MatchedMaxPriority}^{=}\mathsf{\text{-}Condition}(e)=\mathsf{true} & \mbox{ iff $p\in \textit{matched-priorities}(e)$ for a maximal priority} \\
%&\hspace{4mm}\mbox{$p\in priorities(e)$, $\neg\textit{occurOnce}(p,e)$} 
%\end{align*}} 
We give a typical example of an execution $e$ which is not $\mathsf{MatchedMaxPriority}^>$-linearizable in 
%Given a data-differentiated execution $e$ such that $\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Condition}(e)$ holds, Lemma \ref{lemma:automata for extended priority queue with single priority} is not enough for ensure linearizable w.r.t $\mathsf{MatchedMaxPriority}^{>}$. This is because that it is possible that $\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Conc}(e,x)$ does not holds because of interaction between actions of multiple priorities. One such example is shown in 
\figurename~\ref{fig:introduce gap for EPQ1Lar}. Intuitively, this is a violation because during the whole execution of $\textit{rm}(b)$, the priority queue stores a smaller priority value (which should be removed before $b$). To be more precise, we define \emph{the interval of a value $x$} as the time interval from the return of a put $\textit{ret}(\textit{put},x,p)$ to the call of the matching remove $\textit{call}(rm,x)$, or to the end of the execution if such a call action doesn't exist. Intuitively, it represents the time interval in which a value is guaranteed to be stored into the concurrent priority queue. 
In \figurename~\ref{fig:introduce gap for EPQ1Lar}, we draw the interval of each value by dashed line. Here we assume that $p_1 \prec p_4$, $p_2 \prec p_4$ and $p_3 \prec p_4$. We can not find a sequence $s$ where $e \sqsubseteq s$ and $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,b)$ holds, since each time point from $\textit{call}(\textit{rm},b)$ to $\textit{ret}(\textit{rm},b)$ is included in the interval of some smaller priority value, and $\textit{rm}(b)$ can not happen in interval of a smaller priority value.

\begin{figure}[htbp]
TODO UPDATE THE ARGUMENTS OF ACTIONS
  \centering
  \includegraphics[width=0.4 \textwidth]{figures/PIC-HIS-INTRO-GAP-EPQ1L.pdf}
%\vspace{-10pt}
  \caption{An execution that does not linearizable w.r.t $\mathsf{MatchedMaxPriority}^{>}$}
  \label{fig:introduce gap for EPQ1Lar}
\end{figure}

TODO INTRODUCE THE LEFT-RIGHT CONSTRAINT

@@

TODO GIVE THE AUTOMATON

More details can be found in Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}. TODO MOVE THIS TO THE RELATED WORK The notion of left-right constraint used for priority queue is inspired by left-right constraint of queue \cite{Bouajjani:2015}.
